Community Data Science meets Digital Heritage: Controversies about World Heritage Sites in the Online Community of Wikipedia

### 2020-05-21 analysis/paper/paper.Rmd
intro and background

```{r, setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/"
)
```

@@

@@

>>>>

```{r, setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/"
)

# slides: https://drive.google.com/drive/u/1/folders/1TI7JpQHZpMFqxJvNwUPSJhoVCduarCcm
# OSF: https://doi.org/10.17605/OSF.IO/AY27G 
# Video: https://youtu.be/2sWpw8MIDDA 
```

@@

Heritage is the processes and outcomes of people engaging with elements of the past – material and immaterial – and attributing social and cultural meanings to them in the present (Smith 2006; Harrison 2013). These are important to understand because they shape peoples' identities and influence how they think and behave towards other people. Digital heritage are these engagements that are enabled by the Internet [@bonacchi2019digital], leaving traces that can be identified and quantified using data science methods. Digital heritage studies represent a major turn from traditional heritage studies, characterized by post-modernism [@kristiansen2014towards], critical theory, and qualitative methods, towards novel ontologies, data-intensive ethnographies, and a new role for heritage scholars as data scientists. Bonacchi et al. [-@Krzyzanska_2018; -@bonacchi2019digital] have sketched out the new digital heritage research program with their combination of data-intensive and qualitative investigations of 1.4m Facebook posts in Brexit-related community groups. They found recurring parallels – both pro- and anti-Brexit – made by Facebook users between the European Union, the Roman Empire and "barbarians" as they use heritage to support their political activism. They demonstrate the potential for understanding public perceptions and experiences of the past in contemporary society using big data obtained from social media. In this paper we extend the digital heritage research program in two important new directions. First, we introduce Wikipedia as an example of an online peer production community where people engage with elements of the past in measurable ways. Second, present a case study using data science methods to investigate the ways people create and consume Wikipedia articles on cultural sites inscribed on the UNESCO World Heritage List (hereafter CS-WHL).

While social media is a vast and diverse online space that we are only just beginning to explore how people engage with the past, there are other contexts of online interactions where heritage is practiced in distinctive, if poorly understood, ways. We can contrast social media, with its fundamental elements of identity, conversations, sharing, presence, relationships, reputation, and groups [@kietzmann2011social], with peer production communities, where online users participate in the collaborative, asynchronous creation, sharing, promoting, and classification of content in highly structured ways [@wilkinson2008strong]. Online peer production communities are comparable to more traditions kinds of voluntary associations where groups set and execute goals, with explicitly democratic organizational ideals. While the ideals of many online peer production communities emphasize nonhierarchical and nonbureaucratic organization, analysis of large amounts of user activity indicates that most of these communities are actually undemocratic and noninclusive, functioning as entrenched oligarchies [@shaw2014laboratories]. This emphasis on governance and management of collective action is a key detail that distinguishes online communities on social media from online communities on peer production projects. It follows that user interactions in the process of generating context in online peer production communities includes technological and social mechanisms that enact the community's governance policies, for example, limiting a user's activity according to their status in the community's hierarchy, or managing conflict with highly structured procedures. These distinctive organizational and technical qualities of online peer production communities make them an intriguing novel context of heritage production to study digital traces of human activity resulting from engagement with the past

Here we present a study of how people engage with elements of the past in one of the largest and long-lived online peer production communities, Wikipedia, a highly influential (11 billion page views per month. https://stats.wikimedia.org/#/en.wikipedia.org) and well-known online encyclopedia that anyone can edit. Although anyone can edit, most internet users do not, and the characteristics of people who do edit are important to understand to identify biases in the encyclopedia's content. In a survey of 1,512 American adults, @shaw2018pipeline found that majority of internet users (96.5%) had heard of Wikipedia, and 82.5% of those said that they had ever visited the site. Just over two thirds (67.3%) knew that Wikipedia can be edited, but only 8.2% of the full sample had ever edited any page. Factors that strongly predict if a user has ever edited Wikipedia include their gender (male), age (younger), education level (has BA), Internet use frequency (higher) and Internet use skills (higher). @hill2013wikipedia a similarly found that among U.S. adult contributors to the English-language edition of Wikipedia, at least 75% are male. While gender disparities have been documented throughout Wikipedia [@adams2015wikipedia; @ford2017anyone], there are also geographical disparities. @johnson2016not examined 218,709 English geotagged articles about places in the
contiguous United States and 46,124 Chinese geotagged articles about places in China to investigate differences in articles about urban and rural places. They found that articles about rural areas have systematically lower quality, are less likely to have been produced by contributors who focus on the local  area, and are more likely to have been generated by bots (automated software agents).

The software that Wikipedia runs on, MediaWiki, is a complex toolkit that structures user behaviors in ways that are well suited to data science methods the enable efficient computational analysis of large numbers Wikipedia articles. While Wikipedia has elements that are ubiquitous on the Internet, such as links on articles that take the user to other articles or other pages on the Internet, it also has several less common elements that contribute to its unique technicity. The technical schemas of MediaWiki result in very specific types of relationships between human users and the technical elements of the Wikipedia project. For example, every edit to every article is tracked in a publicly accessible version control system. This exposes the article creation process in highly granular detail; we can see how many editors contributed, the size of their edits and their distribution over time. Wikipedia has a special category of edit called the 'revert' which allows a user to restore an article to an earlier state to remove recent vandalism (such as the addition of irrelevant or offensive material).

Technicity: The efficacy, functionality, or experience of a particular technology. relationship between technology and humanity (or humans), Simondon‘s philosophy, technicity is undertaken as a result of the relationship among technical objects and humans.
An incredible sociotechnical system of content production and maintenance with human and non-human actors: people, bots: https://www.makeuseof.com/tag/bots-edit-wikipedia-humans-made/  (eg. rambot) and “software-assisted human editors” https://journals.sagepub.com/doi/pdf/10.1177/1461444810365297 (good figs here)
Editing mechanics (http://www.contropedia.net/ ), bots (fighting https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0171774 but see https://dl.acm.org/citation.cfm?id=3134684), reverts, Talk pages
Conflict detection: https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0038869

Specifically we investigate how people interact with Wikipedia articles about cultural sites inscribed on the WHL. We chose the CS-WHL as a bounded set of cultural heritage elements with near-global distribution, broad public interest, a wide temporal distribution in both the age of the cultural sites, and the age of inscription on the WHL, and the high intensity of cultural and political discussions that often surround the inscription of site on the WHL. These qualities make it an ideal data set as an entry point for digital heritage studies of online peer production communities, which are typically goal-driven (e.g. 'write quality articles') compared to social media activity where activity is more event-driven analysis (e.g. 'share reactions to Brexit').

We survey the basic characteristics of content (article length, number of wikilinks out to other pages, number of citations to non-Wikipedia items), consumption (page view counts, wikilinks in from other Wikipedia pages), and production (edit counts, edit densities, edit sizes, number of unique editors per article, talk page length, talk page topics) to answer the question: how does engagement with the past compare on Wikipedia compare to engagement with other topics? We use a random sample of 10,000 Wikipedia articles as a random sample to compare with articles about cultural sites on the UNESCO World Heritage List. We investigate spatial patterns of behaviour on these Wikipedia articles to answer the question:  are the editors of CS-WHL located near the sites they edit, indicating local community interest in the online representation of their heritage, or in other countries, impling a digital colonalism of world heritage information on Wikipedia where one community's heritage is interpreted and communicated by another? Finally we investigate temporal patterns of activity on CS-WHL Wikipedia pages to answer the question of how correlated editing activity is with events outside of Wikipedia relating to the CS-WHL sites, such as their inscription on the WHL.

In this paper we contribute to digital heritage studies into a new online domain

@@

Kristiansen, K. 2014. “Towards a New Paradigm? the Third Science Revolution and Its Possible Consequences in Archaeology.” Current Swedish Archaeology 22: 11–34.
